---
title: 'UK Biobank Metabolite-Disease Association Atlas: Interactive Visualizations'
author: "Tommi Suvitaival, Steno Diabetes Center Copenhagen, tommi.raimo.leo.suvitaival@regionh.dk"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
    fig_width: 9
    fig_height: 9
    dev: jpeg
    toc: yes
  github_document:
    fig_width: 9
    fig_height: 9
    dev: jpeg
    toc: yes
always_allow_html: true
---

# Introduction

* This is a tutorial to creating interactive visualizations of associations between metabolites and diseases.

* The data used in this tutorial have been published in the following publication:

Julkunen, H., Cichońska, A., Tiainen, M. *et al*. **Atlas of plasma NMR biomarkers for health and disease in 118,461 individuals from the UK Biobank**. *Nat Commun* 14, 604 (2023). https://doi.org/10.1038/s41467-023-36231-7

* The data have been downloaded from https://biomarker-atlas.nightingale.cloud/ under the tab *Summary statistics* (accessed 2023-02-20).

* View this document at https://tommi-s.com/Metabolite-Disease-Atlas/ to show all output correctly.

# Preparations

## Load Data

```{r}

data <- 
  readr::read_csv(
    file = "C:/Users/tsuv0001/Documents/data-nonsensitive-local-copies/ukb_nightingale_biomarker_disease_association_atlas.csv" # Path to the data file downloaded from https://biomarker-atlas.nightingale.cloud/ .
  )

```

## List Measured Biomarkers

```{r}

names.biomarkers <- sort( names( table( data$"biomarker_name" ) ) )

names.biomarkers

```

## Omit Lipoprotein Particle Descriptors

```{r}

names.biomarkers.printed <- names.biomarkers

names.biomarkers.printed <-
  names.biomarkers.printed[ 
    !grepl(
      x = names.biomarkers.printed,
      pattern = "DL"
    )
  ]

names.biomarkers.printed

```

## Select Population and Endpoint

```{r}

table( data$"age_group" )

table( data$"endpoint_type" )

```

```{r}

data.plot <- data

data.plot <- data.plot[ data.plot$"age_group" == "Full population", ]

data.plot <- data.plot[ data.plot$"endpoint_type" == "incident", ]

data.plot <- 
  data.plot[ data.plot$"biomarker_name" %in% names.biomarkers.printed, ]

```

# Principal Component Analysis (PCA)

## Preparations

### Create a Wide Data Matrix

* Biomarkers as columns
* Conditions as rows
* Additional columns on condition name and disease group for use in the plot

```{r}

data.plot.pca <-
  tidyr::pivot_wider(
    data = data.plot,
    id_cols = icd10_desc,
    names_from = biomarker_name,
    # values_from = estimate.significant
    values_from = estimate
  )

tmp <- data.plot.pca$"icd10_desc"

# Extract disease group from disease code (condition).
# Extract most increased and most decreased metabolites for each condition.

data.plot.pca <- 
  data.frame( 
    Condition_Name = unlist( data.plot.pca[ , 1 ] ),
    Disease_Group =
      stringr::str_sub(
        string = unlist( data.plot.pca[ , 1 ] ),
        start = 1,
        end = 1
      ),
    Top_Increase =
      colnames( data.plot.pca )[
        apply(
          X = data.plot.pca[ , -1 ],
          MAR = 1,
          FUN = which.max
        ) + 1
      ],
    Top_Decrease =
      colnames( data.plot.pca )[
        apply(
          X = data.plot.pca[ , -1 ],
          MAR = 1,
          FUN = which.min
        ) + 1
      ],
    Condition = NA,
    data.plot.pca[ , -1 ]
    )

# Create a text for tooltip.

data.plot.pca$"Condition" <-
  paste0(
    data.plot.pca$"Condition_Name",
    "\n\nMost Increased: ",
    data.plot.pca$"Top_Increase",
    "\nMost Decreased: ",
    data.plot.pca$"Top_Decrease"
  )

# Set disease codes as rownames.

rownames( data.plot.pca ) <- tmp

```

### Omit Rows with No Values

* Disease codes that have missing data on any of the biomarkers.

```{r}

has.any.missing <-
  apply(
    X = is.na( data.plot.pca[ , -( 1:5 ) ] ),
    MAR = 1,
    FUN = any
  )

table( has.any.missing )

data.plot.pca <- data.plot.pca[ !has.any.missing, ]

```

### Compute PCA

* Omitting the metadata columns.

```{r}

result.pca <- prcomp( data.plot.pca[ , -( 1:5 ) ] )

```

### Show Non-Interactive PCA

* Conditions as points (scores)
* Metabolites as loadings
* Color by disease group
* Further information about the condition in tooltip (via shape; later on used by *plotly::ggplotly*)

```{r}

library( "ggfortify" )

plot.pca <- 
  autoplot(
    result.pca,
    data = data.plot.pca,
    alpha = 0.5,
    loadings.colour = "black",
    loadings.label = TRUE,
    loadings.label.colour = "black",
    shape = "Condition",
    colour = "Disease_Group"
  ) +
  ggplot2::theme( legend.position = "none" )

```

```{r PCA-Non-Interactive}

print( plot.pca )

```

### Create the Interactive PCA Figure

```{r}

plot.pca.interactive <-
  plotly::ggplotly( 
    p = plot.pca,
    tooltip = "shape"
  )

```

## Show the Interactive PCA Figure

```{r PCA-Interactive}

plot.pca.interactive

```

# Heatmap

## Preparations

### Show Non-Significant Values as Zero

```{r}

data.plot$"Association" <- data.plot$"estimate"

data.plot[ data.plot$"pvalue" > 0.01, "Association" ] <- 0

```

### Create a Wide Data Matrix

```{r}

data.plot.wide <-
  tidyr::pivot_wider(
    data = data.plot,
    id_cols = icd10_desc,
    names_from = biomarker_name,
    values_from = Association
  )

tmp <- data.plot.wide$"icd10_desc"

data.plot.wide <- as.matrix( data.plot.wide[ , -1 ] )

rownames( data.plot.wide ) <- tmp

```

### Fill in Missing Disease-Metabolite Values as Zero

```{r}

data.plot.wide[ is.na( data.plot.wide ) ] <- 0

```


### Omit Conditions with No Associations

```{r}

is.all.zeros <- 
  apply(
    X = data.plot.wide == 0,
    MAR = 1,
    FUN = all
  )

table( is.all.zeros )

conditions.all.zeros <- rownames( data.plot.wide )[ is.all.zeros ]

data.plot.wide <- data.plot.wide[ !is.all.zeros, ]

```

### Create Non-Interactive Heatmap

```{r Heatmap-Non-Interactive, fig.width = 48, fig.height = 72 }

result.heatmap <-
  gplots::heatmap.2(
    x = data.plot.wide,
    trace = "none",
    tracecol = "black",
    col = gplots::bluered( n = 99 ),
    margins = c( 15, 15 ),
    srtCol = 45,
    cexCol = 2
  )

```

### Define the Row and Column Order in Long Data

* Use the order from hierarchical clustering of the non-interactive heatmap.
* The should be possible to specify in the call to *plotly::layout*, but for some reason it does not work for the row order.
* Instead, add a running three-digit number as suffix in the condition name to enforce the desired row order based on alphabetical order.

```{r}

data.plot.heatmap <- data.plot

data.plot.heatmap <- 
  data.plot.heatmap[ 
    !( data.plot.heatmap$"icd10_desc" %in% conditions.all.zeros ),
  ]

data.plot.heatmap$"biomarker_name" <-
  factor(
    x = data.plot.heatmap$"biomarker_name",
    levels = colnames( data.plot.wide )[ result.heatmap$"colInd" ]
  )

data.plot.heatmap$"icd10_desc" <-
  factor(
    x = data.plot.heatmap$"icd10_desc",
    levels = rownames( data.plot.wide )[ result.heatmap$"rowInd" ],
    labels = 
      paste( 
        formatC(
          x = 1:nrow( data.plot.wide ),
          flag = "0",
          width = 3
        ),
        rownames( data.plot.wide )[ result.heatmap$"rowInd" ]
      )
  )

```

### Define Tooltip

* Show the p-value of the association between the condition and metabolite.

```{r}

data.plot.heatmap$"annotation" <-
  paste0(
    "p: ",
    data.plot.heatmap$"pvalue"
  )

```

### Set Colorscale Midpoint to White and Zero

```{r}

val.min <- round( min( data.plot.heatmap$"Association" ) * 100 )

val.max <- round( max( data.plot.heatmap$"Association" ) * 100 )

range <- max( abs( val.max ), abs( val.min ) ) * 2

palette <- gplots::bluered( n = range )

palette <- palette[ round( range / 2 + val.min ) : round( range / 2 + val.max ) ]

```

### Show Non-Significant Values as Missing Data

```{r}

data.plot.heatmap$"Association"[ data.plot.heatmap$"pvalue" > 0.01 ] <- NA

```

### Cut Condition Names Shorter

```{r}

data.plot.heatmap$"icd10_desc" <-
  stringr::str_sub(
    string = data.plot.heatmap$"icd10_desc",
    start = 1,
    end = 30
  )

```

### Create the Interactive Heatmap

```{r}

# Create the heatmap.

plot.heatmap.interactive <-
  plotly::plot_ly(
    data = data.plot.heatmap,
    x = ~biomarker_name,
    y = ~icd10_desc,
    z = ~Association,
    text = ~annotation,
    colors = gplots::bluered( n = 99 ),
    type = "heatmap"
  )

# Define the colorscale as symmetric.

plot.heatmap.interactive <-
plotly::colorbar(
  p = plot.heatmap.interactive,
    limits = 
      c( -1, 1 ) * max( x = abs( data.plot.heatmap$"Association" ), na.rm = TRUE )
  )

# Omit axis titles, adjust text size and omit background grid.

plot.heatmap.interactive <-
  plotly::layout(
    p = plot.heatmap.interactive,
    xaxis = 
      list(
        # categoryorder = "array",
        # categoryarray = colnames( data.plot.wide )[ result.heatmap$"colInd"],
        showgrid = FALSE,
        tickfont = list( size = 7 ),
        title = ""
      ),
    yaxis = 
      list(
        # categoryorder = "array",
        # categoryarray = rownames( data.plot.wide )[ result.heatmap$"rowInd" ],
        showgrid = FALSE,
        tickfont = list( size = 7 ),
        title = ""
      )
  )

```

## Show the Interactive Heatmap

```{r Heatmap-Interactive}

plot.heatmap.interactive

```

# SessionInfo

```{r}

utils::sessionInfo()

```


# Appendix

* Copy README.html to index.html to show this document on Github Pages.

```{r index-html}

if ( file.exists( "README.html" ) ) {
  
  file.copy( 
    from = "README.html",
    to = "index.html",
    overwrite = TRUE
  )
  
}
